---
title: "🧬 Spatial Transcriptomics Analysis (part 1)"
format: html
---

# Welcome to the NGS-CN Summer School on Spatial Transcriptomics Analysis!

Spatial transcriptomics allows us to measure gene expression across tissue sections while preserving spatial information, helping us understand not just **what** genes are active, but **where** they are active.

In this practical session, we will explore how to analyze spatial transcriptomics data using **R** and **RStudio**. Whether you are new to spatial data or already familiar with single-cell workflows, this guide will walk you through the essential steps.

Feel free to use your personal workbook to make notes, track your progress, and revisit it after this workshop.

---

In this **first part**, we will explore the dataset, learn how to load it into R, and examine how the data is structured within a Seurat object. We will also perform some initial visualisations to familiarise ourselves with the dataset and its spatial features.

## Setting up your R environment 

Before diving into the data, we need to load a few important R packages. These libraries will provide the tools we need for analysis and visualization.

```{r}
#| label: setup
#| message: false

library(Seurat)     # Main package for single-cell and spatial transcriptomics analysis
library(SeuratWrappers) # Add-ons for Seurat
library(patchwork)  # Helps combine multiple ggplot2 plots into one figure
library(tidyverse)  # Essential collection of R packages, including ggplot2, dplyr, magrittr
library(grid)       # For other plots
library(viridis)    # Our continuous colour palette

# Colours
cal_pal50 = c("#Fa1a8e", "#009B7D", "#ff9933", "#7083a4", "#ffce45", "#015e05", 
              "#fedf25", "#d2b48c", "#bb55e1", "#6ec5ab", "#5d57af", "#143341", 
              "#761445", "#d65b5a", "#94043a", "#e7a6cd", "#204519", "#87afd1", 
              "#9b9a9d", "#f95b70", "#83c874", "#808080", "#452b5b", "#ecb100", 
              "#f46124", "#525252", "#4c84a3", "#00bfff", "#01b4c6", "#174d79", 
              "#a6a0f2", "#76facc", "#8491be", "#a32a2f", "#1c8859", "#2cc012", 
              "#35782b", "#9c6755", "#3b3960", "#eeb0a1", "#3e1e13", "#0064c3", 
              "#d81e4a", "#74646c", "#f675da", "#ffce45", "#ec7014", "#e50000", 
              "#000000", "#a4527c", "#041859")   

# Set default colours for ggplot2
options(ggplot2.discrete.colour=cal_pal50)
options(ggplot2.discrete.fill=cal_pal50)
options(ggplot2.continuous.colour="viridis")
options(ggplot2.continuous.fill="viridis")
options(ggplot2.binned.colour="viridis")
options(ggplot2.binned.fill="viridis")
```

# Our dataset

In this section, we will introduce the dataset we are using for this training session.

## Dataset overview

We use a **mouse brain dataset** from [10x Genomics](https://www.10xgenomics.com/):   

* Source: Downloaded from [here](https://www.10xgenomics.com/datasets/visium-hd-cytassist-gene-expression-libraries-of-mouse-brain-he)   
* Sample: Male C57BL/6 mouse, 8 weeks old   
* Tissue: Brain tissue block (Formalin-Fixed Paraffin-Embedded, or FFPE)   
* Platform: Visium HD Spatial Gene Expression kit   
* Sequencing: Illumina NovaSeq 6000, approximately 300 million reads   
* Processing: Space Ranger was already run to process the raw data   

The dataset is located in the `datasets/visiumhd_mouse_brain` directory within your project folder. If you need to download the dataset again, you can use the `datasets/download.R` script.

## Dataset files

The dataset includes several files and folders:

* `web_summary.html`: Quality control (QC) report in HTML format
* `metrics_summary.csv`: QC report in CSV format
* `cloupe_008um.cloupe`: Loupe file for visualizing the data in the Loupe Browser
* `molecule_info.h5`: H5 file with information about reads and molecules in the dataset
* `feature_slice.h5`: H5 file for high-performance data access of 2 µm bins
* `spatial`: Folder containing the tissue images
* `binned_outputs` Folder containing results organized by different bin sizes
   * `square_002um/`: Binned data with 2 µm resolution
   * `square_008um/`: Binned data with 8 µm resolution
   * `square_0016um/`: Binned data with 16 µm resolution

The folder `binned_outputs` is relevant for our analysis. Space Ranger defaults to grouping the data into square bins of 2 µm, 8 µm, and 16 µm, though these bin sizes can be modified in the Space Ranger settings.

## Focusing on `square_008um`

Let us focus on the `square_008um/` folder, which contains the binned results at 8 µm resolution. This folder includes:

* `analysis/`: Results from Space Ranger analysis
* `cloupe.cloupe`: Loupe file for visualizing the 8 µm binned data in Loupe Browser
* `raw_feature_bc_matrix/`: Raw counts for all bins (tissue or no tissue)
* `raw_feature_bc_matrix.h5`: As above, organized in H5 format
* `filtered_feature_bc_matrix/`: Counts for all bins located under tissue, per bin and gene
* `filtered_feature_bc_matrix.h5`: As above, organized in H5 format
* `spatial/`:  Folder with images and information about the spatial locations of the bins on the tissue

The key files for our analysis are located in `filtered_feature_bc_matrix`. 

## Counts data

Let us take a closer look at the `filtered_feature_bc_matrix/` directory, which includes three important files:

* `barcodes.tsv.gz`: Contains the names (or identifiers) for each of the bins in the dataset
* `features.tsv.gz` Contains essential information about the genes, including their Ensembl IDs, symbols and gene types
* `matrix.mtx.gz` Contains the actual counts data
   * The first row contains the total number of genes, bins and counts.
   * All other rows contain the gene index (row in `features.tsv.gz`), bin index (row in `barcodes.tsv.gz`) and the count value.
   * Only non-zero counts are stored.

Here are the first few lines of each file for the 016 µm dataset:

```{r}
#| label: dataset_1

# barcodes.tsv.gz
read.delim("datasets/visiumhd_mouse_brain/binned_outputs/square_016um/filtered_feature_bc_matrix/barcodes.tsv.gz", 
           header=FALSE, nrows=5)

# features.tsv.gz
read.delim("datasets/visiumhd_mouse_brain/binned_outputs/square_016um/filtered_feature_bc_matrix/features.tsv.gz", 
           header=FALSE, nrows=5)

# matrix.mtx.gz
read.delim("datasets/visiumhd_mouse_brain/binned_outputs/square_016um/filtered_feature_bc_matrix/matrix.mtx.gz", 
           skip=2, nrows=5, header=FALSE, sep=" ")
```

::: {.callout-tip}
### Example
This dataset has 19,059 genes, 98,917 bins and 68,259,814 counts.  
Row 2: Gene #**17** (Vxn) has in bin #**1** (s_016um_00052_00082-1) a count of **3**.
:::

::: {.callout-note}
### Assignment
Find out how many bins, genes and counts are in the 8µm dataset. 
:::

# The Seurat object

The R package [Seurat](https://satijalab.org/seurat/) is a widely used tool for single-cell and spatial transcriptomics analysis. It allows researchers to work with complex biological data and perform tasks like dimensionality reduction, clustering, and visualization. 

Another popular option is the [scanpy](https://squidpy.readthedocs.io/en/stable/) package in Python. 

## Importing the data into Seurat

The first step in analyzing your data is to import it into Seurat. Here, we load data for both 8µm and 16µm bins into a Seurat object: 

```{r}
#| label: dataset_2
sc = Load10X_Spatial("datasets/visiumhd_mouse_brain", bin.size=c(8, 16), assay="RNA")
sc$orig.ident = "brain"
Idents(sc) = "brain"
```

Now, let us take a look at the Seurat object to understand its structure:

```{r}
#| label: dataset_3

sc
```

::: {.callout-note}
### Questions
How many assays are in the dataset?  
How many images are in the dataset?
Why always two?
:::

## What is stored in the Seurat object?

The Seurat object organizes data into different components. These components facilitate spatial transcriptomics analysis and are crucial for data interpretation. The key components include:

- **Bins and genes**: These represent the spatial units (bins) and the genes that are being measured
- **Assay data**: Contains the gene expression data, chromatin accessibility, surface proteins, etc.
- **Metadata for bins**: Information about each bin (e.g., sample type, condition)
- **Metadata for genes**: Includes gene IDs, names, types, and descriptions
- **Dimensionality reductions**: Methods like PCA, UMAP, and tSNE that are used to reduce the complexity of the data
- **Neighbors and graphs**: Information about the relatedness between bins
- **Images**: Spatial coordinates and image data for visualizing the tissue

## Bins and genes

We can retrieve the total number of genes and bins stored in the Seurat object like this:

```{r}
#| label: dataset_4

dim(sc)
```

We can also list the bins and genes contained in the Seurat object:

```{r}
#| label: dataset_5

Cells(sc) %>% head(5)
Features(sc) %>% head(5)
```

::: {.callout-tip}
### Nomenclature
Note that sometimes the term **bin** is also referred to as barcode, cell, nuclei.   
The term **gene** is also referred to as feature, protein, peak.   
We will stick to bins and genes for this workshop. 
:::

Furthermore, bins have a default identity which is used for plotting. It can be accessed with the `Idents` function:

```{r}
#| label: dataset_6

Idents(sc) %>% head(5)
```

## Assay data

Assays store the actual data, for example gene expression counts. We can list the available assays in the Seurat object:

```{r}
#| label: dataset_7

Assays(sc)
```

To access a specific assay, we use the `GetAssay` function or the `[[]]` syntax:

```{r}
#| label: dataset_8

GetAssay(sc, assay="RNA.008um")
sc[["RNA.008um"]]
```

We can also check the number of genes and bins in a specific assay like this:

```{r}
#| label: dataset_9

assay = GetAssay(sc, assay="RNA.008um")
dim(assay)
```

::: {.callout-note}
### Questions
How many bins are in the assay?   
How many genes are in the assay?
:::

The assay object stores data in different **layers**:  

* `counts`: Raw counts for each gene in each bin  
* `data`: Normalized counts for better comparability across bins  
* `scale.data`: Scaled data for better comparability across genes  

We use the `GetAssayData` function to access the actual data:  

```{r}
#| label: dataset_10
#| message: true

counts = GetAssayData(sc, assay="RNA.008um", layer="counts")
counts[1:10, 1:10]
```

::: {.callout-note}
### Questions
What do you find special about this matrix?
:::

The Seurat object has a default/active assay that is used for most functions. To check the active assay, we can use:

```{r}
#| label: dataset_11

DefaultAssay(sc)
```

To change the active assay to another one, for example `"RNA.016um"`, use:

```{r}
#| label: dataset_12

DefaultAssay(sc) = "RNA.016um"
DefaultAssay(sc)
```

For this workshop, we will focus on the assay `"RNA.008um"` for our analysis:

```{r}
#| label: dataset_13

DefaultAssay(sc) = "RNA.008um"
DefaultAssay(sc)
```

::: {.callout-tip}
### Default assay
Most Seurat functions will automatically use the default assay, but you can always specify a different assay with the `assay=` argument.
:::

## Bin metadata

Bin metadata store additional information about each bin, such as the sample or condition. You can access this metadata using the `[[]]` syntax:

```{r}
#| label: dataset_14

sc[[]] %>% head(5)
```

For example, the `orig.ident` column stores the original identity of the bin, and the `nCount_RNA.008um` and `nFeature_RNA.008um` columns store the number of counts and genes for the 8µm bins.

You can access individual metadata columns like this:

```{r}
#| label: dataset_15

sc$nCount_RNA.008um %>% head(3)
```

## Gene metadata

Gene metadata can be accessed using the `[[]]` syntax on the Seurat assay object:

```{r}
#| label: dataset_16
#| message: true

assay = GetAssay(sc, assay="RNA.008um")
assay[[]] %>% head(5)
```

In this example, no gene metadata is available for the assay. If available, you can access individual columns using the `$` syntax.

## Dimensionality reductions

Dimensionality reductions (e.g., PCA, UMAP, tSNE) are essential for visualizing high-dimensional data in a lower-dimensional space. You can list available reductions with the `Reductions` function:

```{r}
#| label: dataset_17

Reductions(sc)
```

Currently, there are no reductions in the Seurat object. If they were present, you could access them using the `Reductions` function or the `[[]]` syntax.

## Neighbors and graphs

To identify how the bins are related (e.g., for clustering), Seurat computes neighbors and graphs. You can list available neighbors and graphs using:

```{r}
#| label: dataset_18

Neighbors(sc)
Graphs(sc)
```

Currently, there are no computed neighbors or graphs. If available, you could access them with the `GetNeighbors` / `GetGraph` functions or the `[[]]` syntax.

## Images

For spatial data, images are stored in the Seurat object to visualize tissue sections and spatial coordinates. To list available images, use:

```{r}
#| label: dataset_19

Images(sc)
```

Each assay and bin size should have a corresponding image. You can access a specific image using the `[[]]` syntax:

```{r}
#| label: dataset_20

sc[["slice1.008um"]]
```

Note that the image also has a default assay meaning the image is linked to the assay. Whenever we plot data from this assay in a spatial context, this image will be used.

### Tissue image

When we refer to the image in spatial transcriptomics, we usually mean the spatial coordinates associated with each bin. These coordinates tell us where each bin is located within the tissue section but not what the tissue looks like visually.

However, Seurat also allows us to view the underlying histological image that was used to derive these spatial coordinates. This can help us visually inspect the tissue morphology in parallel with gene expression patterns.

In this example, **Hematoxylin and eosin (H&E) staining** was used. H&E staining is a widely used technique in histology and pathology for visualizing the morphological features of tissues. Hematoxylin stains cell nuclei a blue to purple hue by binding to nucleic acids, highlighting areas rich in DNA and RNA. Eosin is an acidic dye that stains the cytoplasm and extracellular matrix in varying shades of pink to red, depending on the protein content and tissue composition. The contrast between the basophilic (blue-purple) and eosinophilic (pink-red) structures enables clear differentiation of cellular components and tissue architecture, aiding in diagnostic assessment and research.

Here is how we can visualize the raw tissue image grid along with the spatial:

```{r}
#| label: dataset_22
#| out-width: 4in
#| out-height: 4in
#| fig-align: center

gridExtra::grid.arrange(GetImage(sc[["slice1.008um"]], method="grob"))
```

By default, Seurat loads the low-resolution version of the image (``tissue_lowres_image.png` in Spaceranger output). See here - https://github.com/satijalab/seurat/issues/8996 - how load the high-resolution version of the image (`tissue_hires_image.png` in Spaceranger output) instead. 

### Coordinates

Spatial coordinates are available in the Seurat in three different coordinate systems depending on the underlying image:

- **original**: Pixel coordinates for the original, full resolution image (provided by the user when running Spaceranger)
- **hires**: Pixel coordinates for a downsampled, high-resolution version of the original image (`tissue_hires_image.png` in Spaceranger output)
- **lowres**: Pixel coordinates for a downsampled, low-resolution version of the original (`tissue_lowres_image.png` in Spaceranger output)



and can be accessed using the `GetTissueCoordinates` function. This function retrieves the coordinates of the bins on the tissue image:


```{r}
#| label: dataset_21

coordinates = GetTissueCoordinates(sc, image="slice1.008um")
coordinates %>% head(5)
```


### Image coordinates
We can get the image coordinates of the bins (in pixel) using the `GetTissueCoordinates` function by setting the `scale` parameter to the `lowres` image:

```{r}
#| label: dataset_23

coordinates = GetTissueCoordinates(sc, image="slice1.008um", scale="lowres")
coordinates %>% head(5)
```

The maximum of `x` is approximately in image width in pixel. The maximum of `y` is approximately in image height in pixel.


### Spatial coordinates and image coordinates

Now we have learnt that there are spatial coordinates and an (underlying) image. This means that there are two coordinate systems here:

- **spatial coordinates**: The coordinates of the bins on the Visium HD slide. These vary depending on bin size.
- **image coordinates**: The coordinates of the pixels in the image. These vary depending on the image resolution.

This is illustrated in the following (a bit complicated) plot where we plot one bin in its image coordinates and label it adding its spatial coordinates:

```{r}
#| label: dataset_24

# Select a bin
bin = "s_008um_00301_00321-1"

# Get image and spatial coordinates
spatial_coordinates = GetTissueCoordinates(sc, image="slice1.008um")
bin_spatial_x = round(coordinates[bin, "x"])
bin_spatial_y = round(coordinates[bin, "y"])

image_coordinates = GetTissueCoordinates(sc, image="slice1.008um", scale="lowres")
bin_image_x = image_coordinates[bin, "x"]
bin_image_y = image_coordinates[bin, "y"]

# Make a spatial plot
p = SpatialDimPlot(sc, cells.highlight=bin)

# Add the image coordinates to the plot, remove legend and relabel axes
p = p + theme_minimal() + NoLegend()
p = p + xlab("image_x") + ylab("image_y")

# Highlight the bin bigger
p = p + 
  annotate("point", x=bin_image_x, y=bin_image_y, size=1, color="#FFFF00") +
  annotate("text", x=bin_image_x, y=bin_image_y, label=paste0("spatial_x=", bin_spatial_x, " spatial_y=", bin_spatial_y), vjust=-1)

# Add bin info as title
p = p + ggtitle(bin)
p
```

Spatial coordinates are converted into image coordinates by multiplying them with scale factor. The scale factors can be accessed using `ScaleFactors` function on the image object. 

```{r}
#| label: dataset_25

# Get spatial coordinates
spatial_coordinates = GetTissueCoordinates(sc, image="slice1.008um") %>% head()

# Get the scale factors that convert to image coordinates
scale_factors = ScaleFactors(sc[["slice1.008um"]])[["lowres"]]

# Convert to image coordinates
image_coordinates = spatial_coordinates
image_coordinates$x = spatial_coordinates$x * scale_factors
image_coordinates$y = spatial_coordinates$y * scale_factors
image_coordinates
```

When we compare the converted output to the output of the `GetTissueCoordinates` function with `scale="lowres"`, we can see that they are the same:

```{r}
#| label: dataset_26

# Compare to coordinates from function
GetTissueCoordinates(sc, image="slice1.008um", scale="lowres") %>% head()
```

### Zooming into image regions

We can also zoom into specific regions of the image. To do so, we provide coordinates or a list of bins and Seurat will create a "new image" (a field of view that crops the original image).

First, we can use the `Crop` function to select a specific, rectagular region:

```{r}
#| label: dataset_27

# Using tissue/spatial coordinates (x: 5000 - 8000, y: 5000 - 8000)
sc[["region1"]] = Crop(sc[["slice1.008um"]], x=c(5000, 8000), y=c(5000, 8000))

p1 = SpatialFeaturePlot(sc, images="region1", "nCount_RNA.008um", pt.size.factor = 10)

# Using image coordinates (x: 100 - 200, y: 100 - 200)
scale_factors = ScaleFactors(sc[["slice1.008um"]])[["lowres"]]
sc[["region2"]] = Crop(sc[["slice1.008um"]], x=c(100, 120) / scale_factors, y=c(100, 120) / scale_factors, scale="lowres")

p2 = SpatialFeaturePlot(sc, images="region2", "nCount_RNA.008um", pt.size.factor = 20)

# Using a list of bins. We simply select all bins using image coordinates.
selected_bins = GetTissueCoordinates(sc, image="slice1.008um", scale="lowres") %>%
  filter(x > 300 & x < 340 & y > 300 & y < 340) %>%
  pull(cell)

sc[["region3"]] = subset(sc[["slice1.008um"]], cells=selected_bins)
p3 = SpatialFeaturePlot(sc, images="region3", "nCount_RNA.008um", pt.size.factor = 10)


p1 | p2 | p3
```

For more complicated (non-rectangular) regions, we first create a segmentation (a polygon) that is our selection mask. In a real analysis, you could use Loupe Browser or Xenium Explorer or another image software to do this.

```{r}
#| label: dataset_28

# A shaped selection (we start from the top left corner)
segmentation_mask = data.frame(
  x = c(8000, 10000, 10000, 9000, 8000, 8000),
  y = c(8000, 8000, 6000, 7000, 6000, 8000)
)
segmentation_mask$cell = "selection1"
segmentation_mask = CreateSegmentation(segmentation_mask)
```

We then use the `Overlay` function to select the region:

```{r}
#| label: dataset_29

# Note: there is also an invert argument here
sc[["region4"]] = Overlay(sc[["slice1.008um"]], segmentation_mask)

p1 = SpatialFeaturePlot(sc, images="region4", "nCount_RNA.008um", pt.size.factor = 10)
```

**Important**: We now have four additional images that are all associated with the `RNA.008um` assay. This means that whenever we plot data from this assay in a spatial context, all images will be plotted. 

To avoid this behaviour, we can explictly specify an image when plotting. Alternatively, we set the default assay for the additional images to an empty string (meaning they will not be included anymore).

```{r}
#| label: dataset_28

DefaultAssay(sc[["region1"]]) = ""
DefaultAssay(sc[["region2"]]) = ""
DefaultAssay(sc[["region3"]]) = ""
DefaultAssay(sc[["region4"]]) = ""
```


## Plotting

Now that we have the data in Seurat, we can start visualizing gene expression using plots.

### Violin plot for gene expression

We can create a **violin plot** to show the expression of specific genes (e.g., "Nrgn", "Igf2", "Hap1") in the RNA.008um assay:

```{r}
#| label: dataset_29

VlnPlot(sc, features=c("Nrgn", "Igf2", "Hap1"), layer="counts", pt.size=0) + NoLegend()
```

Since the counts are raw and not normalized, interpreting them directly can be challenging.

### Spatial plot for gene expression

We can also visualize gene expression in a spatial context using a **spatial feature plot**:

```{r}
#| label: dataset_30
#| dpi: 75

SpatialFeaturePlot(sc, features=c("Nrgn", "Igf2", "Hap1"), slot="counts")
```

::: {.callout-tip}
### Plot size and resolution matter
You may observe visual artefacts in the image. This can occur when the image is plotted at a relatively low resolution (e.g., 150 dpi) and subsequently viewed on a high-resolution display. To address this issue, consider increasing the resolution using the Quarto `dpi` option. Alternatively, reducing the figure size may also help mitigate the problem.
:::

### Comparison Between 8µm and 16µm Bins

Finally, we can plot the gene expression in the **RNA.016um assay**. This allows us to visualize how the data changes when comparing different bin sizes.

```{r}
#| label: dataset_31

DefaultAssay(sc) = "RNA.016um"
p1 = VlnPlot(sc, features=c("Nrgn"), layer="counts", pt.size=0) + NoLegend()
p2 = SpatialFeaturePlot(sc, features=c("Nrgn"), slot="counts")
DefaultAssay(sc) = "RNA.008um"
p1 | p2
```

::: {.callout-note}
### Assignment
Make a violin plot and a spatial feature plot for the gene **Hap1** for the **8µm assay**. Make a violin plot and a spatial feature plot for the gene **Prkcd** for the **16µm assay**. 
:::

::: {.callout-note}
### (Bonus-)Assignment
Explore the structure of the Seurat object with the `str()` function.
:::

# Session info

To enhance reproducibility and facilitate the sharing of your analysis, it is good practice to include information about the R session and the packages used:

```{r}
#| label: session_info

sessioninfo::session_info()
```

**Now that we have successfully loaded the data into Seurat and explored the structure of the Seurat object, it's time to move on to the next step: data analysis.**

**To free the occupied memory, please restart the R session in RStudio via Session->Restart R.**